Vagueness, Noise, and Signaling

### 2014-11-05 paper/paper_short.tex
paper shortened, and plots resized/rearranged

\begin{abstract}
  Signaling games are appealing models for studying the evolution of
  meaning, but typically do not explain vagueness as a feature of
  successful signaling.  Complementing recent like-minded models, we
  introduce the replicator diffusion dynamic as a special case of the
  replicator mutator dynamic that combines evolutionary pressure on
  optimal signal use with stochastic noise in the processing of
  similar stimuli. Applying this new dynamic to a generalization of
  Lewis' signaling games, we show that stochastic imprecision leads to
  vague, yet communicative efficient signal use, and, moreover, also
  unifies evolutionary outcomes and helps avoid suboptimal
  categorization.
\end{abstract}

Many of our concepts and words are vague. A vague category knows clear
cases that fall under it, clear cases that do not, and also so-called
borderline cases. Borderline cases do not clearly apply, nor do they
clearly not apply, and there may be differences between borderline
cases in terms of how well they represent the category in
question. Vagueness does not seem to dramatically affect the success
of everyday communication, but it is troublesome for some of the most
prominent theories of language and meaning. This is especially the
case for the logico-semantic tradition of Frege, Russell and the young
Wittgenstein which is challenged by the paradoxes vagueness gives rise
to. But there are other intriguing aspects about vagueness.

The puzzle that we are concerned with here is how vagueness could
arise and be maintained in the first place. This is a serious worry to
functionalist accounts that maintain that our concepts and language
use evolved in order to be efficient. Since the existence of unclear
borderline cases seems to entail inefficiency of categorization or
communication, the challenge, succinctly put forward by
\citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
exist despite its obvious and demonstrable functional deficiency under
evolutionary pressure to be optimally expressive. We address
\emph{Lipman's problem} as a technical challenge,
namely to find a conceptually sound and mathematically coherent formal
model of the evolution of categorization and language use that yields
by-and-large efficient categories \emph{and} vagueness.

A number of authors have recently tried to explain why vagueness
evolved as something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of
limitations in information processing
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This
paper contributes to the latter line of thought. Concretely, we
introduce the replicator diffusion dynamics ---a novel variant of the
replicator mutator dynamic--- that integrates stochastic noise on the
differential confusability of similar stimuli. The dynamic proposed
here generalizes and complements previous like-minded accounts. We
show that stochastic noise can not only lead to vague, yet
communicative efficient signal use, but can also unify evolutionary
outcomes and help avoid suboptimal categorization.

@@

A sim-max game, in the relevant sense here, consists of a set of
states $\States$, a set of messages $\Messgs$ with much fever messages
than states, a prior probability distribution $\Pr \in
\Delta(\States)$ that gives the occurrence probabilities of states, a
similarity metric on states $\similarity \mycolon \States \times
\States \rightarrow \mathds{R}$, and a utility function $\Utils
\mycolon \States \times \States \rightarrow \mathds{R}$. We identify
the receiver's acts with the states of the world, so that the game is
one of guessing the actual state, so to speak. We also assume that
sender's and receiver's interests are alike, so we only have one
utility function. We do not consider message costs, so utilities only
depend on the actual state, and the receiver's response. The
similarity function should satisfy the usual requirements for a metric
and the utility function should be a monotonically decreasing function
of similarity.

\citet{JagerMetzger2011:Voronoi-Languag} showed that the
evolutionarily stable states of sim-max games are remarkably
systematic. Their results were obtained for games with infinitely many
states in $n$-dimensional Euclidean space $\States \subseteq
\mathds{R}^n$ and a quadratic loss function for utilities
$\Utils(\mystate{1}, \mystate{2}) = - (\mystate{1} -
\mystate{2})^2$. For these games, the evolutionarily stable states are
demonstrably so-called Voronoi languages. Roughly put, a Voronoi
language is a pair of sender and receiver strategies, such that the
sender strategy partitions the state space into convex categories,
while the receiver's interpretations are the central spots in each
category. For example, if \States is the unit interval and all states
are equiprobable, an \ESS of a sim-max game with two messages would
have the sender use one message exclusively for all points in the
lower half of the unit interval and another for all points in the upper
half; the receiver's interpretations of messages are the central
points in the respective intervals.

This result is interesting, because it demonstrates that signaling can
impose orderly categories on a metric space, without that being the
ulterior purpose of it all. Sender and receiver can be distinct
entities, whose purpose is to communicate effectively about the actual
state. In that case, evolving Voronoi languages would explain why
\emph{linguistic categories} are well-behaved and orderly in the way
they appear to be. More abstractly, sender and receiver can also be
thought of as distinct modules in a single system, where the first
module must discretize the information by selecting a small sample of,
suggestively, category labels. These are passed to a second module
that tries to decode the original information. In this case, evolving
Voronoi languages would explain why \emph{conceptual categories} are
well-behaved and orderly in the way that they appear to be.

Seen in this light, sim-max games may provide a foundation to
approaches in cognitive semantics that rely on the notion of
conceptual spaces.  \citet[][70--77]{Gardenfors2000:Conceptual-Spac},
for example, has prominently argued that natural categories are convex
regions in conceptual space. If the conceptual space has a suitable
metric, convex categories can be derived from a set of prototypes. The
category corresponding to prototype $p$ is the set of points that are
more similar to $p$ than to any other
\citep[e.g.][]{OkabeBoots2000:Spatial-Tessell}. In this way,
\citet{Gardenfors2000:Conceptual-Spac} argues, an efficient
categorization system can be obtained: storing the prototypes lets us
recover the categories without having to store each category's
extension. However, what is left unexplained, is where the prototypes
come from, and why we would not see just any distribution of
prototypes as an equally efficient classification system. This is
where sim-max games can contribute a principled approach to deriving,
in an independent way, not only convex categories but also
prototypical exemplars belonging to them.

@@

This outline of an approach to categorization using sim-max games
leaves many problems unaddressed. One of them is that natural
categories for continuously variable stimuli usually do not have
unique, point-valued prototypes and clear category boundaries. We
would like to account for the possibility of such vagueness, in
particular for the following criteria: (i) clear positive examples of
a vague category should show a gradient transition to clear negative
examples to (so-called \emph{higher-order vagueness}); (ii) prototypes
should likewise be gradient regions, peaking at the center of the
vague category they represent
\citep[e.g.][]{Sainsbury1991:Is-There-Higher,KeefeSmith1997:Vagueness:-A-Re}.

\citet{DouvenDecock2011:Vagueness:-A-Co} show that
\citeauthor{Gardenfors2000:Conceptual-Spac}'s conceptual spaces
approach can be extended to account for the existence of borderline
cases. From the assumption that prototypes are extended, yet convex
regions in conceptual space, they give a construction algorithm that
yields ``collated Voronoi diagrams'' with thick boundaries
representing borderline
regions. \citet{DecockDouven2012:What-is-Graded-} show further how it
is possible to arrive at higher-order vagueness, by weighing in the
distance of different borderline cases to various prototypical
regions. This accounts for the first of the two desiderata mentioned
above, but still assumes that crisp prototype regions must be
given.

@@

    \begin{tikzpicture}[node distance = 2cm, thick]

@@

      \node[rectangle, draw=mycol, fill=mycol!25, thick] (actual)
      {actual state $\mystate{i}$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      actual] (perceived) {perceived state $\mystate{j}$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      perceived] (output) {chosen message $\messg$};

      \node[rectangle, thick, below of = output, node distance=1cm]
      (sender) {sender};

@@

      \node[rectangle, draw=mycol, fill=mycol!25, thick] (message)
      {observed message $\messg$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      message] (target) {target interpretation $\mystate{i}$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      target] (realized) {realized interpretation $\mystate{j}$};

      \node[rectangle, thick, below of = realized, node distance=1cm]
      (sender) {receiver};

@@

      \node[below of = mixedSpace, node distance = 1.7cm] {space of
        mixed strategies};
    \end{scope}

@@

      \node[below of = behSpace, node distance = 1.7cm] {space of
        behavioral strategies};
    \end{scope}

@@

For example, consider a signaling game with two states and two
messages. Let the confusion matrix be:
\begin{align*}
  C=
  \begin{pmatrix}
    .8 & .2 \\
    .2 & .8 
  \end{pmatrix}\,.
\end{align*}
The resulting mutation matrices are:
\begin{align*}
Q^C & = \bordermatrix{ ~ & 11 & 12 & 21 & 22 \cr
                      11 & 1 & 0 & 0 & 0 \cr
                      12 & .16 & .64 & .04 & .16 \cr
                      21 & .16 & .04 & .64 & .16 \cr
                      22 & 0 & 0 & 0 & 1 \cr}
&                    
  R^C & = \bordermatrix{ ~ & 11 & 12 & 21 & 22 \cr
                      11 & .64 & .16 & .16 & .04 \cr
                      12 & .16 & .64 & .04 & .16 \cr
                      21 & .16 & .04 & .64 & .16 \cr
                      22 & .04 & .12 & .12 & .64 \cr}\,.
\end{align*}
Here, a pair like $21$, for example, refers to a pure sender strategy
with $\Spure(\state_1)=\messg_2$ and $\Spure(\state_2) =
\messg_1$. Similarly for the receiver.

Based on the translations from confusion matrices $C$ to mutation
matrices $Q^C$ and $R^C$ in Equations~(\ref{eq:construction-rec}) and
(\ref{eq:construction-sen}), we can finally formulate the
desired correspondence result, that reveals the \rdd as a special case
of the \rmd in behavioral strategies where the only source of mutation is
imperfect discriminability of states.

@@

\noindent A proof is given in Appendix~\ref{sec:proofs}.

@@

  \includegraphics[width=0.85\textwidth]{plots/NosofskySim.pdf}

@@

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_NS-90_tol-01_imp0_ind3098.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0$}
    \label{fig:example_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_NS-90_tol-01_imp01_ind3452.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0.1$}
    \label{fig:example_stratsB}
  \end{subfigure}

@@

To further explore our simulation results, we defined a small number
of metrics that aim to capture numerically how vague, communicatively
efficient and generally well-structured the resulting strategy pairs
were. \emph{Entropy} and \emph{informativity} capture (slightly
different aspects of) the amount of systematicity or regularity in
signal use. \emph{Voronoiness} and \emph{convexity} are, respectively,
gradient and categorical measures for the coherence of signal-induced
categories. \emph{Expected utility} measures the communicative
efficiency of evolved strategy pairs.

@@

The sender strategies in Figures~\ref{fig:example_stratsA} and
\ref{fig:example_stratsB} have entropy $1.19e^{-5}$ and $0.21$,
respectively. The receiver strategies have respective entropies $0.43$
and $0.84$. In general, we expect that vague languages will have
higher entropy than crisp ones and that increasing imprecision will
also increase entropy, all else being equal.

\paragraph{Informativity.} The quantity of information in a signal is
an old notion that also goes back to the start of information theory.
\citet[ch.~3]{Skyrms2010:Signals} discusses its use in the context of
signaling games. Intuitively speaking, informativity in Skyrms' sense
measures how revealing the sender's signal use is about which state is
actual. It therefore measures, in our given set-up with flat priors, a
sense of well-behavedness or regularity of a sender strategy.

The main idea behind Skyrms' notion is that we can quantify the amount
of information about a state $\state$ in a message $\messg$ via the
relation between the probability of the state given the message
$\Pr(\state \,|\, \messg)$ and the unconditional probability of the
state $\Pr(\state)$.  Following Bayes' theorem, we can express
$\Pr(\state \,|\, \messg)$ as $\frac{\Pr(\state) \cdot \Pr(\messg
  \,|\, \state)}{\Pr(\messg)}$.  We then have $\Pr(\messg \,|\,
\state) = \Sstrat(\state,\messg)$ and $\Pr(\messg) =
\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot
\Sstrat(\state^\prime,\messg)$.  Finally, we equate sender
infomativity with the average overall information about states in each
signal.  Following \citet[p.~36]{Skyrms2010:Signals}, and the
considerations above, the informativity about states conveyed by a
sender strategy is:
\begin{align*}
  I(\Sstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs}
  \sum_{\state \in \States} \frac{\Pr(\state) \cdot
    \Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States}
    \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \cdot \log
  \left(\frac{\Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States}
      \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \right)
  \,. 
\end{align*}
This measure is bounded between $0$ and $1$. The sender strategies in
Figures~\ref{fig:example_stratsA} and \ref{fig:example_stratsB} have
informativity $0.69$ and $0.54$, respectively. In general, we expect
that vague languages will have lower informativity than crisp
ones and that increasing imprecision will decrease informativity.

\citet[p.~39]{Skyrms2010:Signals} also offers a notion of
informativity about the acts performed by the receiver, but it turns
out that this is perfectly correlated with information about the
states in the results of our simulations. We therefore only look at
informativity about states.

% Conversely, we can quantify the amount of information about an act in a message.
% We equate receiver informativity with the average overall information about acts in each signal.
% Based on Skyrms' definition~\citet[p.~39]{Skyrms2010:Signals}, we define receiver informativity as follows:
% \begin{align*}
%   I(\Sstrat,\Rstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs} \sum_{\act \in \Acts} \Rstrat(\messg,\act) \cdot \log \left(\frac{\Rstrat(\messg,\act)}{\sum_{\state \in \States} \Pr(\state) \cdot \sum_{\messg^\prime \in \Messgs} \Sstrat(\state,\messg^\prime) \cdot \Rstrat(\messg^\prime,\act)} \right) \,.
% \end{align*}

\paragraph{Voronoiness.}
This metric aims to quantify how close a strategy is to being a part
of a Voronoi tessallation of the state space. Bearing in mind that
this metric is specifically geared towards the sim-max games in our
set-up, we define Voronoiness as follows. Let $\pi(\Rstrat, \messg) =
\argmax_{\state \in \States}\Rstrat(\messg,\state)$ be the prototype
of a message $\messg \in \Messgs$, then
\begin{align*}
  V(\Sstrat, \Rstrat) & = \sum_{\state \in \States} \sum_{\messg_1 \in
    \Messgs} \sum_{\messg_2 \in \Messgs} F(\Sstrat, \Rstrat, \state,
  \messg_1, \messg_2) \text{\ \ \ with } \\
  F(\Sstrat, \Rstrat, \state,
  \messg_1, \messg_2) & =
  \begin{cases}
    1 & \text{if } \similarity(\state, \pi(\Rstrat, \messg_1)) >
    \similarity(\state, \pi(\Rstrat, \messg_2)) \bicond \\
    &  \ \ \ \ \ \Sstrat(\state, \messg_1) >
    \Sstrat(\state, \messg_2) \\
    0 & \text{otherwise.}
  \end{cases}
\end{align*}
is the Voronoiness of the sender strategy $\Sstrat$ given the receiver
strategy $\Rstrat$ and
\begin{align*}
  V(\Rstrat) & = \sum_{\messg \in
    \Messgs} \sum_{\state_1 \in \States}  \sum_{\state_2 \in \States}
  F(\Rstrat, \messg, \state_1, \state_2) \text{\ \ \ with } \\
  F(\Rstrat, \messg, \state_1, \state_2) & =
  \begin{cases}
    1 & \text{if } \similarity(\state_1, \pi(\Rstrat,
  \messg)) > \similarity(\state_2, \pi(\Rstrat, \messg)) \\
  & \ \ \ \ \ \bicond \Rstrat(\messg,
  \state_1) > \Rstrat(\messg, \state_2) \\
    0 & \text{otherwise.}
  \end{cases}
\end{align*}
is the Voronoiness of the receiver strategy $\Rstrat$.  The metrics
are lower bound by $0$ and upper bound by, respectively, $|\States|
\times |\Messgs|^2$ and $|\Messgs| \times |\States|^2$, thus we will
normalize them by dividing by these values.

@@

Figure~\ref{fig:MeanMetrics} shows the means of the recorded metrics,
together with their estimated 95\% confidence intervals, for each
triple of independent variables. Most trends are quite clear, and not
unexpected. Entropy of sender and receiver strategies are increasing
with imprecision. Entropy of sender strategies seems independent of
tolerance levels and the number of states, while there seems to be an
effect of both parameters on the entropy of the receiver
strategy. Sender informativity and expected utility are decreasing
under growing imprecision. There seem to be mild effects of the number
of states in the game, but hardly any in terms of pragmatic
tolerance. What was not expected \emph{a priori} was that Voronoiness
of both sender and receiver strategies appear to be non-decreasing
under increasing impairment, with the only exception being the 6 states
games and the transition from $\impairment = 0$ to $\impairment =
0.05$. Expected utility of evolved languages mostly decreases in
increasing imprecision, with the sole exception of a rise for $\toler
= 0.2$ and $n_s = 6$ at $\impairment = 0.1$. Still, normalized expected
utility of evolved strategy pairs under mild values of imprecision are
not necessarily much worse than their respective perfect precision
\rd-analogues. This is especially so for lower numbers of states
(which increase the chance of non-convergence and non-convexity; see
Figure~\ref{fig:CategoricalMeasures} and the discussion below).

Taken together, these findings suggest the following effect of
confusability of states and diffusion of behavior. Mild forms of
perceptual imprecision lead to slightly less communicative efficiency,
more vagueness and more regular, well-behaved languages. This
impression is also supported by the proportion of converged and convex
trials, pictured in
Figure~\ref{fig:CategoricalMeasures}. Non-converged and non-convex
sender strategies showed for several values of parameters in the
absence of impairment. Increasing impairment always assured
convergence and convexity: diffusion speeds up convergence to a convex
and regular language.

  \includegraphics[width=0.8\textwidth]{plots/CategoricalMeasures.pdf}

  \caption{Proportion of converged and convex trials.}
  \label{fig:CategoricalMeasures}
\end{figure}

@@

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_NS-6_tol-03_imp005_ind41.pdf}
    \caption{$\ns = 6$, $\toler = 0.2$, $\impairment = 0$}
    \label{fig:example_stratsC}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_NS-6_tol-03_imp005_ind23.pdf}
    \caption{$\ns = 6$, $\toler = 0.2$, $\impairment = 0$}
    \label{fig:example_stratsD}
  \end{subfigure}

@@

Diffusion, in this context, could be either of two things. It could be
differential mutation probabilities in line with other interpretations
of mutation in the \rmd
\citep[e.g.][]{NowakKomarova2001:Evolution-of-Un,KomarovaNiyogi2001:The-Evolutionar}:
pure strategies are not faithfully reproduced, e.g. by imperfections
of learning or underdetermination of input, and mistakes in inheriting
pure strategies are more likely if they result from the confusion of
similar states. Another possible biological interpretation of the
\rdd, is that inheritance is faithful, but strategies are noisily
realized. Strategies are not necessarily selected for what they are,
but rather for how they are realized, once noise is factored in.

@@

Under this non-biological interpretation, we have again two options of
picturing what diffusion is, similar to the biological cases
before. One possibility is that diffusion is noise in the adoption of
strategies, e.g. by conditional imitation, of strategies by other
signalers. Another possibility is, again, that diffusion is noise on
the realization of strategies: while behavior is optimized to be
efficient (be it due to learning, introspection or imitation),
realization of strategies is bound to be noisy due to confusability of
similar states.

>>>>

\begin{abstract}
  Signaling games are popular models for studying the evolution of
  meaning, but typical approaches do not allow for vagueness as a
  feature of successful signaling.  Complementing recent like-minded
  models, we introduce the replicator diffusion dynamic as a special
  case of the replicator mutator dynamic that combines evolutionary
  pressure on optimal signal use with stochastic noise in the
  processing of similar stimuli. Applying this new dynamic to a
  generalization of Lewis' signaling games, we show that stochastic
  imprecision leads to vague, yet communicative efficient signal use,
  and, moreover, also unifies evolutionary outcomes and helps avoid
  suboptimal categorization.
\end{abstract}

Many of our concepts and words are vague. A vague category knows clear
cases that fall under it, clear cases that do not, and also so-called
borderline cases. Borderline cases do not clearly apply, nor do they
clearly not apply, and there may be differences between borderline
cases in terms of how well they represent the category in
question. Vagueness does not seem to dramatically affect the success
of everyday communication, but it is troublesome for some of the most
prominent theories of language and meaning. This is especially the
case for the logico-semantic tradition of Frege, Russell and the young
Wittgenstein which is challenged by the paradoxes vagueness gives rise
to.

But there are other intriguing aspects about vagueness. The puzzle
that we are concerned with here is how vagueness could arise and be
maintained in the first place. This is a serious worry to
functionalist accounts that maintain that our concepts and language
use evolved in order to be efficient. Since the existence of unclear
borderline cases seems to entail inefficiency of categorization or
communication, the challenge, succinctly put forward by
\citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
exist despite its obvious and demonstrable functional deficiency under
evolutionary pressure to be optimally expressive. We address
\emph{Lipman's problem} as a technical challenge, namely to find a
conceptually sound and mathematically coherent formal model of the
evolution of categorization and language use that yields by-and-large
efficient categories \emph{and} vagueness.

A number of authors have recently tried to explain why vagueness
evolved as something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of
limitations in information processing
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This
paper contributes to the latter line of thought. Concretely, we
introduce the replicator diffusion dynamics ---a novel variant of the
replicator mutator dynamic--- that integrates stochastic noise on the
differential confusability of similar stimuli. The dynamic proposed
here generalizes and complements previous like-minded accounts. We
show that stochastic noise can not only lead to vague, yet efficient
signal use, but can also unify evolutionary outcomes and help avoid
suboptimal categorization.

@@

A sim-max game, in the relevant sense here, consists of a set of
states $\States$, a set of messages $\Messgs$ with much fever messages
than states, a prior probability distribution $\Pr \in
\Delta(\States)$ that gives the occurrence probabilities of states, a
similarity metric on states $\similarity \mycolon \States \times
\States \rightarrow \mathds{R}$, and a utility function $\Utils
\mycolon \States \times \States \rightarrow \mathds{R}$. We identify
the receiver's acts with the states of the world, so that the game is
one of guessing the actual state, so to speak. We also assume that
sender's and receiver's interests are alike, so we only have one
utility function. We do not consider message costs, so utilities only
depend on the actual state and the receiver's response. The
similarity function should satisfy the usual requirements for a metric
and the utility function should be a monotonically decreasing function
of similarity.

\citet{JagerMetzger2011:Voronoi-Languag} showed that the
evolutionarily stable states of sim-max games are remarkably
systematic. Their results were obtained for games with infinitely many
states in $n$-dimensional Euclidean space $\States \subseteq
\mathds{R}^n$ and a quadratic loss function for utilities
$\Utils(\mystate{1}, \mystate{2}) = - (\mystate{1} -
\mystate{2})^2$. For these games, the evolutionarily stable states are
demonstrably so-called Voronoi languages. Roughly put, a Voronoi
language is a pair of sender and receiver strategies, such that the
sender strategy partitions the state space into convex categories,
while the receiver's interpretations are the central spots in each
category. For example, if \States is the unit interval and all states
are equiprobable, a Voronoi language with two messages would have the
sender use one message exclusively for all points in the lower half of
the unit interval and another for all points in the upper half; the
receiver's interpretations of messages are the central points, .25 and
.75, in the respective intervals.

This result is interesting, because it demonstrates that signaling can
impose orderly categories on a metric space, without that being the
ulterior purpose of it all. Sender and receiver can be distinct
entities, whose purpose is to communicate effectively about the actual
state. In that case, evolving Voronoi languages would explain why
\emph{linguistic categories} are well-behaved and orderly in the way
they appear to be. More abstractly, sender and receiver can also be
thought of as distinct modules in a single system, where the first
module must discretize the information it is fed by selecting a small
sample of, suggestively, category labels. These are passed to a second
module that tries to decode the original information. In this case,
evolving Voronoi languages would explain why \emph{conceptual
  categories} are well-behaved and orderly in the way that they appear
to be.

Seen in this light, sim-max games may provide a foundation to
approaches in cognitive semantics that rely on the notion of
conceptual spaces.  \citet[][70--77]{Gardenfors2000:Conceptual-Spac},
for example, has prominently argued that natural categories are convex
regions in conceptual space. If the conceptual space has a suitable
metric, convex categories can be derived from a set of prototypes. The
category corresponding to prototype $p$ is the set of points that are
more similar to $p$ than to any other. In this way,
\citet{Gardenfors2000:Conceptual-Spac} argues, an efficient
categorization system can be obtained: storing the prototypes lets us
recover the categories without having to store each category's
extension. However, what is left unexplained, is where the prototypes
come from, and why we would not see just any distribution of
prototypes as an equally efficient classification system. This is
where sim-max games can contribute a principled approach to deriving,
in an independent way, not only convex categories but also
prototypical exemplars belonging to them.

@@

This outline of an approach to categorization using sim-max games
leaves many problems unaddressed. One of them is that natural
categories for continuously variable stimuli usually do not have
unique, point-valued prototypes and clear category boundaries. We
would like to account for the possibility of such vagueness, in
particular: (i) clear positive examples of
a vague category should show a gradient transition to clear negative
examples (so-called \emph{higher-order vagueness}); (ii) prototypes
should likewise be gradient regions, peaking at the center of the
vague category they represent
\citep[e.g.][]{Sainsbury1991:Is-There-Higher,KeefeSmith1997:Vagueness:-A-Re}.

\citet{DouvenDecock2011:Vagueness:-A-Co} show that
\citeauthor{Gardenfors2000:Conceptual-Spac}'s conceptual spaces
approach can be extended to account for the existence of borderline
cases. From the assumption that prototypes are extended, yet convex
regions in conceptual space, a construction algorithm is available that
yields ``collated Voronoi diagrams'' with thick boundaries
representing borderline
regions. \citet{DecockDouven2012:What-is-Graded-} show further how it
is possible to arrive at higher-order vagueness, by weighing in the
distance of different borderline cases to various prototypical
regions. This accounts for the first of the two desiderata mentioned
above, but still assumes that crisp prototype regions must be
given.

@@

    \begin{tikzpicture}[node distance = 1.6cm, thick]

@@

      \node[rectangle, draw=black!50, fill=black!10, thick] (actual)
      {actual state $\mystate{i}$};

      \node[rectangle, draw=black!50, fill=black!10, thick, below of =
      actual] (perceived) {perceived state $\mystate{j}$};

      \node[rectangle, draw=black!50, fill=black!10, thick, below of =
      perceived] (output) {chosen message $\messg$};

      \node[rectangle, thick, above of = actual, node distance=0.75cm]
      (sender) {sender};

@@

      \node[rectangle, draw=black!50, fill=black!10, thick] (message)
      {observed message $\messg$};

      \node[rectangle, draw=black!50, fill=black!10, thick, below of =
      message] (target) {target interpretation $\mystate{i}$};

      \node[rectangle, draw=black!50, fill=black!10, thick, below of =
      target] (realized) {realized interpretation $\mystate{j}$};

      \node[rectangle, thick, above of = message, node distance=0.75cm]
      (receiver) {receiver};

@@

      \node[above of = mixedSpace, node distance = 1.4cm] {space of
        mixed strategies};
    \end{scope}

@@

      \node[above of = behSpace, node distance = 1.4cm] {space of
        behavioral strategies};
    \end{scope}

@@

Based on Equations~(\ref{eq:construction-rec}) and
(\ref{eq:construction-sen}), we can formulate the desired
correspondence result. A proof is given in Appendix~\ref{sec:proofs}.

@@

@@

  \includegraphics[width=0.7\textwidth]{plots/NosofskySim.pdf}

@@

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind3098.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0$}
    \label{fig:example_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind3452.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0.1$}
    \label{fig:example_stratsB}
  \end{subfigure}

@@

To further explore our simulation results, we defined a small number
of metrics that aim to capture numerically how vague, communicatively
efficient and generally well-structured the resulting strategy pairs
were. \emph{Entropy} captures the amount of systematicity or
regularity in signal use. \emph{Voronoiness} and \emph{convexity} are,
respectively, gradient and categorical measures for the coherence of
signal-induced categories. \emph{Expected utility} measures the
communicative efficiency of evolved strategy pairs.

@@

The sender strategies in Figures~\ref{fig:example_stratsA} and
\ref{fig:example_stratsB} have entropy $1.19e^{-5}$ and $0.21$,
respectively. The receiver strategies have respective entropies $0.43$
and $0.84$. In general, we expect that vague languages will have
higher entropy than crisp ones and that increasing imprecision will
increase entropy, all else equal.

\paragraph{Voronoiness.}
This metric aims to quantify how close a strategy is to being a part
of a Voronoi tessallation of the state space. Bearing in mind that
this metric is specifically geared towards the sim-max games in our
set-up, we define Voronoiness as follows. Let $\pi(\Rstrat, \messg) =
\argmax_{\state \in \States}\Rstrat(\messg,\state)$ be the prototype
of a message $\messg \in \Messgs$, then
\begin{align*}
  V(\Sstrat, \Rstrat) & = \sum_{\state \in \States} \sum_{\messg_1 \in
    \Messgs} \sum_{\messg_2 \in \Messgs} F(\Sstrat, \Rstrat, \state,
  \messg_1, \messg_2) \text{\ \ \ with } \\
  F(\Sstrat, \Rstrat, \state,
  \messg_1, \messg_2) & =
  \begin{cases}
    1 & \text{if } \similarity(\state, \pi(\Rstrat, \messg_1)) >
    \similarity(\state, \pi(\Rstrat, \messg_2)) \bicond \\
    &  \ \ \ \ \ \Sstrat(\state, \messg_1) >
    \Sstrat(\state, \messg_2) \\
    0 & \text{otherwise}
  \end{cases}
\end{align*}
is the Voronoiness of the sender strategy $\Sstrat$ given the receiver
strategy $\Rstrat$ and
\begin{align*}
  V(\Rstrat) & = \sum_{\messg \in
    \Messgs} \sum_{\state_1 \in \States}  \sum_{\state_2 \in \States}
  F(\Rstrat, \messg, \state_1, \state_2) \text{\ \ \ with } \\
  F(\Rstrat, \messg, \state_1, \state_2) & =
  \begin{cases}
    1 & \text{if } \similarity(\state_1, \pi(\Rstrat,
  \messg)) > \similarity(\state_2, \pi(\Rstrat, \messg)) \\
  & \ \ \ \ \ \bicond \Rstrat(\messg,
  \state_1) > \Rstrat(\messg, \state_2) \\
    0 & \text{otherwise}
  \end{cases}
\end{align*}
is the Voronoiness of the receiver strategy $\Rstrat$.  The metrics
are lower bound by $0$ and upper bound by, respectively, $|\States|
\times |\Messgs|^2$ and $|\Messgs| \times |\States|^2$, thus we will
normalize them by dividing by these values.

@@

Figure~\ref{fig:MeanMetrics} shows the means of the recorded metrics,
together with their estimated 95\% confidence intervals, for each
triple of independent variables. Most trends are quite clear, and not
unexpected. Entropy of sender and receiver strategies are increasing
with imprecision. Entropy of sender strategies seems independent of
tolerance levels and the number of states, while there seems to be an
effect of both parameters on the entropy of the receiver
strategy. What was not expected \emph{a priori} was that Voronoiness
of both sender and receiver strategies appear to be non-decreasing
under increasing impairment, with the only exception being games with
6 states and the transition from $\impairment = 0$ to $\impairment =
0.05$. Expected utility of evolved languages mostly decreases in
increasing imprecision, with the sole exception of a rise for $\toler
= 0.2$ and $n_s = 6$ at $\impairment = 0.1$. Still, normalized
expected utility of evolved strategy pairs under mild values of
imprecision are not necessarily much worse than their respective
perfect precision \rd-analogues. This is especially so for lower
numbers of states (see discussion below).

Taken together, these findings suggest the following effect of
confusability of states and diffusion of behavior. Mild forms of
perceptual imprecision lead to slightly less communicative efficiency,
more vagueness and more regular, well-behaved languages. This
impression is also supported by the proportion of converged and convex
trials, pictured in
Figure~\ref{fig:MeanMetrics}. Non-converged and non-convex
sender strategies showed for several values of parameters in the
absence of impairment. Increasing impairment always assured
convergence and convexity: diffusion speeds up convergence to a convex
and regular language.

% \begin{figure}
%   \centering

%   \includegraphics[width=0.8\textwidth]{plots/CategoricalMeasures.pdf}

%   \caption{Proportion of converged and convex trials.}
%   \label{fig:CategoricalMeasures}
% \end{figure}

@@

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind21.pdf}
    \caption{$\ns = 6$, $\toler = 0.2$, $\impairment = 0$}
    \label{fig:example_stratsC}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_ind23.pdf}
    \caption{$\ns = 6$, $\toler = 0.2$, $\impairment = 0$}
    \label{fig:example_stratsD}
  \end{subfigure}

@@

Diffusion, in this context, could be either of two things. It could be
differential mutation probabilities in line with other interpretations
of mutation in the \rmd
\citep[e.g.][]{NowakKomarova2001:Evolution-of-Un,KomarovaNiyogi2001:The-Evolutionar}:
pure strategies are not faithfully reproduced, e.g. by imperfections
of learning or underdetermination of input, and mistakes in inheriting
pure strategies are more likely if they result from the confusion of
similar states. Another possible biological interpretation of the
\rdd is that inheritance is faithful, but strategies are noisily
realized. Strategies are not necessarily selected for what they are,
but rather for how they are realized, once noise is factored in.

@@

Under this non-biological interpretation, we have again two options of
picturing what diffusion is. One possibility is that diffusion is
noise in the adoption of strategies, e.g. by conditional imitation, of
strategies by other signalers. Another possibility is, again, that
diffusion is noise on the realization of strategies: while behavior is
optimized to be efficient (be it due to learning, introspection or
imitation), realization of strategies is bound to be noisy due to
confusability of similar states.
